### 2013-03-06 paper_sean/results.tex
jmc last edit\!
Final version of paper

The simulator enabled HyperNEAT to evolve fast, natural gaits. In
simulation, HyperNEAT gaits were
significantly faster than those from Glette et al. (\figref{threeplots}-Top, $p <
6.8\times10^{-8}$ when comparing the best gaits in the final
generation of each run via Matlab's Wilcoxon rank sum test).
Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4
cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD). To facilitate comparisons to earlier works~\cite{yos:clune,glette} we report 
mean $\pm$ SD, but our qualitative conclusions are the same when using medians.
Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.

%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
\figOneOnTwo[h!]{threeplots}{avg_fit_vs_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
{
\textbf{Top: }
%HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator. Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines). HyperNEAT gaits are 52.1\% faster in
HyperNEAT outperforms a genetic algorithm with a simple encoding \cite{glette} when both algorithms are combined with a simulator. Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines). HyperNEAT gaits are 52.1\% faster in
simulation and 5.1\% faster in reality than those from a previous study \cite{glette}~(details in \tabref{results}).
\textbf{Left: }
Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced smooth and symmetrical gaits that contained complex regularities. 
\textbf{Right: }Mean gait frequency averaged over 20 runs. 
Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learned to produce gaits with frequencies low enough to avoid this penalty.
}

@@

The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
to a total difference of 40000 vs. 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
potential cause of improved performance is the lower noise in the simulator, which could have helped HyperNEAT find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
substantial, preventing effective learning~\cite{yos:clune}. 
To investigate this hypothesis, we performed 20 runs in simulation with only 180 fitness evaluations, which was the number used in Yosinski et al.~\cite{yos:clune}. The simulated gaits performed
slightly, but not significantly, better than those evolved in hardware ($p = 0.1571$, mean fitness 7.9 cm/s).
Reduced noise thus may have had a small affect on performance, but the substantial performance gains that resulted from using a simulator likely came from the additional evaluations the simulator afforded.
The encouragement of low-frequency gaits in this study
also may have aided performance, especially since in simulation the gaits were
high-frequency in a few early generations before rapidly settling to a
 range below the penalized threshold~(\figref{threeplots}-Right).

While this study was able to produce the fastest QuadraTot gait to
date, most of the gaits in simulation did not transfer well to
reality. Many gaits that were fast in simulation performed
poorly on the real robot, largely due to servos that were too weak and
shut down, or because of differences between simulation and
reality. Repeated attempts to minimize these problems were
unsuccessful. 
%A major, unavoidable cause of the servo shutdowns was
%the design of the robot: the force required to lift the robot in many
%poses was too much for the servos. 
In future studies we will use a
robot that has more mechanical advantage and requires less torque from
each servo, such as the Aracna platform~\cite{lohmann2012aracna}.
%which was designed in response to the challenges experienced with the
%QuadraTot platform. 
That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models. In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal,koos2011transferability}.

>>>>

The simulator enabled HyperNEAT to evolve fast, natural gaits. In
simulation, HyperNEAT gaits were faster than those from Glette et al. (\figref{threeplots}-Top, $p <
6.8\times10^{-8}$ when comparing the best gaits in the final
generation of each run via Matlab's Wilcoxon rank sum test).
Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4
cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD). To facilitate comparisons to earlier works~\cite{yos:clune,glette} we report 
mean $\pm$ SD, but our qualitative conclusions are the same when using medians.
Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.

%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
\figOneOnTwo[h!]{threeplots}{avg_fit_vs_gen}{0.605}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
{
\textbf{Top: }
%HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator. Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines). HyperNEAT gaits are 52.1\% faster in
HyperNEAT outperforms a genetic algorithm with a simple encoding \cite{glette} when both algorithms are combined with a simulator. Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines). HyperNEAT gaits are 52.1\% faster in
simulation and 5.1\% faster in reality than those from a previous study \cite{glette}~(details in \tabref{results}).
\textbf{Left: }
Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced smooth and symmetrical gaits that contained complex regularities. 
\textbf{Right: }Mean gait frequency averaged over 20 runs. 
Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learned to produce gaits with frequencies low enough to avoid this penalty.
}

@@

The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
to a total difference of 40000 vs. 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
potential cause of improved performance is the lower noise in the simulator, which could have helped HyperNEAT find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
substantial, preventing effective learning~\cite{yos:clune}. 
To investigate this hypothesis, we performed 20 runs in simulation with only 180 fitness evaluations, which was the number used in Yosinski et al.~\cite{yos:clune}. The simulated gaits performed
slightly, but not significantly, better than those evolved in hardware ($p = 0.1571$, mean fitness $7.9 \pm 2.14$ cm/s).
Reduced noise thus may have had a small affect on performance, but the substantial performance gains that resulted from using a simulator likely came from the additional evaluations the simulator afforded.
The encouragement of low-frequency gaits in this study
also may have aided performance, especially since in simulation the gaits were
high-frequency in a few early generations before rapidly settling to a
 range below the penalized threshold~(\figref{threeplots}-Right).

While this study was able to produce the fastest QuadraTot gait to
date, most of the gaits in simulation did not transfer well to
reality. Many gaits that were fast in simulation performed
poorly on the real robot, largely due to servos that were too weak and
shut down, or because of differences between simulation and
reality. Repeated attempts to minimize these problems were
unsuccessful. 
%A major, unavoidable cause of the servo shutdowns was
%the design of the robot: the force required to lift the robot in many
%poses was too much for the servos. 
In future studies we will use a
robot that has more mechanical advantage and requires less torque from
each servo, such as the Aracna platform~\cite{lohmann2012aracna}.
%which was designed in response to the challenges experienced with the
%QuadraTot platform. 
That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models. In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{bongard,zagal,koos2011transferability}.
